\thetitle

### 2020-11-14 writing/challenges.tex
edits in intor, challenges and mutate

\section{Why is it hard to evaluate the value of blame?}
\label{sec:challenges}

The most serious challenge for evaluating blame concerns 
the nature of this research. Unlike ordinary
work on programming languages, the question seems to call for empirical research
similar to the one in the human-computer interaction area. At the same time, a
significant evaluation result demands a large number of cases yielding a serious
amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
around this dilemma is to postulate the equivalent of a {\it homo economicus\/}
for programming languages---the {\em rational programmer\/}---and to implement
this rational programmer as an algorithm.  As indicated in the introduction, in
the context of gradual typing, the rational programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion. Measuring this simulated
programmer's behavior on a large number of inputs then yields relevant 
data, similar to collecting empirical data in a human-focused discipline.

A derivative challenge is the problem of simulating the programmer's
mistakes or, more precisely, the origin of impedance mismatches. Before a
programmer can try to find mistakes, mistakes have to be made.
Specifically, a necessary step to test the rational programmer is to find
a collection of  type mistakes in  mix-typed programs that is highly
representantive of both mistakes and programs ``in the wild.''
Unfortunately no such collection exists.  There is a good reason for that;
the kind of mistakes the rational programmer targets are easily detected
by unit tests and, even if their sources are hard to find, these mistakes do not make it
into code repositories with appropriate commit messages.  In other words, the problem is how to generate a
sufficiently large and representantive corpus of type mistakes for a meaningful
experiment.  Following~\citet{lksfd-popl-2020} again, the solution is to
use mutation analysis~\cite{lipton1971fault, demillo1978hints,
jia2011analysis} to create mistakes in a synthetic manner. This process
yields a large number of buggy programs whose mixed-typed versions are
suitable subjects for our experiment.  The mutation analysis of Lazarek et
al, is, however, completely useless for creating impedance mismatches.

@@

The second big challenge is the disparity of gradually
typed languages. While the approaches are well understood in theory,
they come with vastly different type systems, accommodate seemingly
incompatible language features, and apply to radically different coding
styles in the untyped world.  To eliminate the effect of these differences 
on its results, the application of an evaluation method to different
gradual typing systems must use the same test programs and the same type
system. Otherwise, results for different gradual typing systems 
do not share a common baseline and, therefore, do not lend support to any
conclusions about whether the value of blame changes from one system to
another. For instance, the case study presented here builds on
Greenman and Felleisen's transient variant for functional Typed
Racket~\cite{gf-icfp-2018} to obtain a transient semantics that is
comparable with Typed Racket's default natural one. The revised transient
variant accommodates many more linguistic features and library calls
in a way that is faithful to the transient
semantics~\cite{vss-popl-2017}.

>>>>

\section{Why is it hard to evaluate blame?}
\label{sec:challenges}

The most serious challenge for evaluating blame concerns 
the nature of this research. Unlike ordinary
work on programming languages, the question seems to call for empirical research
similar to the one in the human-computer interaction area. At the same time, a
significant evaluation result demands a large number of cases yielding a serious
amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
around this dilemma is to postulate the equivalent of a {\it homo economicus\/}
for programming languages---the {\em rational programmer\/}---and to implement
this rational programmer as an algorithm.  As indicated in the introduction, in
the context of gradual typing, the rational programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion. Measuring this simulated
programmer's behavior on a large number of inputs yields relevant 
data, similar to collecting empirical data in a human-focused discipline.

A derivative challenge is the problem of simulating the programmer's
mistakes or, more precisely, the origin of impedance mismatches. Before a
programmer can try to find mistakes, mistakes have to be made.
Specifically, a necessary step to test the rational programmer is to find
a collection of  type mistakes in  mix-typed programs that is highly
representantive of both mistakes and programs ``in the wild.''
Unfortunately no such collection exists.  There is a good reason for that;
the kind of mistakes the rational programmer targets are easily detected
by unit tests and, even if their sources are hard to find, these mistakes do not make it
into code repositories with appropriate commit messages.  In other words, the problem is how to generate a
sufficiently large and representantive corpus of type mistakes for a meaningful
experiment.  Following~\citet{lksfd-popl-2020} again, the solution is to
use mutation analysis~\cite{lipton1971fault, demillo1978hints,
jia2011analysis} to create mistakes in a synthetic manner. This process
yields a large number of buggy programs whose mixed-typed versions are
suitable subjects for our experiment.  The mutation analysis of Lazarek et
al, is, however, useless for creating impedance mismatches.

@@

The second big challenge is the disparity of gradually
typed languages. While the approaches are well understood in theory,
they come with vastly different type systems, accommodate seemingly
incompatible language features, and apply to radically different coding
styles in the untyped world.  To eliminate the effect of these differences 
on its results, the application of an evaluation method to different
gradual typing systems must use the same test programs and the same type
system. Otherwise, results for different gradual typing systems 
do not share a common baseline and, therefore, do not lend support to any
conclusions about whether the value of blame changes from one system to
another. For instance, the case study presented here builds on
Greenman and Felleisen's transient variant for functional Typed
Racket~\cite{gf-icfp-2018} to obtain a transient semantics that is
comparable with Typed Racket's default natural one. This transient
variant accommodates many more linguistic features and library calls
in a way that is faithful to the transient
semantics~\cite{vss-popl-2017}.

### 2020-11-14 writing/introduction.tex
edits in intor, challenges and mutate

Industrial implementors of  gradual typing systems have almost completely ignored blame assignment.
Systems such as Flow,\footnote{\url{https://flow.org}}
Hack,\footnote{\url{https://hacklang.org}} or
TypeScript\footnote{\url{https://www.typescriptlang.org}} exploit types for IDE actions and for finding typos in
code. Then their compilers remove types and rely on the built-in safety checks of
the underlying programming language to catch any problems. 
Recently, \citet{cc-snapl-19} have turned this implicitly expressed
skepticism into an explicit statement and challenge.

This contrast between theoretical research and industrial applications of gradual typing raises the question 
\begin{quote}
 \it
 whether blame assignment adds any value to a gradually typed language, \\
 especially for the benefit of the working programmer.
\end{quote}
Given the long standing academic interest in blame  and its complete absence in
industrial systems, it comes as an even bigger surprise that the research
literature and the industrial blog world do not present or discuss any
possible answers.  Instead, when language designers decide to look into the
question, they seem to answer it one way or another without any scientific
justification. As a matter of fact, the research community simply does not have a
method for evaluating blame, and given the question's practical
relevance, this a significant problem.

@@

The paper's second contribution is a set of results from applying the evaluation
method to three distinct blame assignment strategies and approximately 20,000
cases\mf{Replace with real number} (section~\ref{sec:results}): Vitousek et
al.'s strategy of tracking typed/untyped boundaries in Reticulated;
Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
mf-toplas-2009}; and the approach of industrial systems that forego blame
and instead rely on error messages from the
underlying programming language.

>>>>

Industrial implementors of  gradual typing systems have almost completely ignored blame assignment.
Systems such as Flow,\footnote{\url{https://flow.org}}
Hack,\footnote{\url{https://hacklang.org}} or
TypeScript\footnote{\url{https://www.typescriptlang.org}} exploit types for IDE actions and for finding typos in
code. Then their compilers remove types and rely on the built-in safety checks of
the underlying language to catch any problems. 
Recently, \citet{cc-snapl-19} have turned this implicitly expressed
skepticism into an explicit statement and challenge.

This contrast between theoretical research and industrial applications of gradual typing raises the question 
\begin{quote}
 \it
 whether blame assignment adds any value to a gradually typed language,
 especially for the benefit of the working programmer.
\end{quote}
Given the long standing academic interest in blame  and its complete absence in
industrial systems, it comes as an even bigger surprise that the research
literature and the industrial blog world do not present or discuss any
possible answers.  Instead, when language designers decide to look into the
question, they seem to answer it one way or another without any scientific
justification. As a matter of fact, the research community simply does not have a
method for evaluating blame, and given the question's practical
relevance, this is a significant problem.

@@

The paper's second contribution is a set of results from applying the evaluation
method to three distinct blame assignment strategies and approximately
65000 cases (section~\ref{sec:results}): Vitousek et
al.'s strategy of tracking typed/untyped boundaries in Reticulated;
Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
mf-toplas-2009}; and the approach of industrial systems that forego blame
and instead rely on error messages from the
underlying programming language.

### 2020-11-14 writing/mutate.tex
edits in intor, challenges and mutate

However,  unlike the deep errors introduced by standard mutators, 
gradual type systems detect shallow errors.  Thus, in order to
evaluate the rational programmer in the context of gradual typing, 
we need to develop alternative mutators
that are fine-tuned to generate shallow errors.  Our inspiration is our
more-than-a-decade long experience of making type-level mistakes in Typed
Racket. Some of these mistakes often take non-trivial effort to debug and
have become the seed for our mutators, which we summarize in
figure~\ref{table:mutation-ops}.  For each one, the figure provides a
short description and an example mutation.

Most of our mutators are self-explanatory.
The first twelve apply to most gradually typed languages, including those with classes.
The last four mutators target distinguishing features of Typed Racket.
Specifically, \texttt{arithmetic} may replace a \texttt{+} with a \texttt{-}
in an attempt to change the type of the result of the arithmetic
operation. In Typed Racket, \texttt{+}'s result is a
\texttt{Positive-Integer} when all arguments are
\texttt{Positive-Integer}. However, the result of \texttt{-} is
\texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
advantage of Typed Racket's ``truthiness.'' Finally, \texttt{negate-cond} and \texttt{force-cond}
aim to confuse occurrence typing.

@@

The astute reader must have observed that mutators that cause interesting
type-level mistakes do not necessarily turn our benchmarks into
interesting debugging scenarios for the rational programmer.  The
benchmarks are fully typed, and type checking flags each impedance at
compile time.  Hence, we apply the mutators to versions, dubbed
\dubbed{configurations}, of the benchmarks where some components are
stripped of their type annotation.  Without the annotations, the type
checker may not detect an impedance and, instead, the type-level checks
of the gradual type system signal a run time error.  Mutant configurations
with type-level mistakes that result in run time errors are the ideal
ground to exercise the rational programmer's ability to locate errors.

Specifically, we define an \emph{interesting debugging scenario} as a
mutant configuration that
\begin{enumerate} 
  \item is rejected by the type checker when we restore any missing type annotations, 
  \item produces a run time error under Erasure as-is, and 
  \item the stack trace of the error mentions at least three distinct components.  
\end{enumerate}

The first criterion simply validates that the debugging scenario concerns
a type-level mistake. However, The second and third criteria require a
little more explanation.  The second one stipulates that the mistake is
detectable under Erasure to ensure that it is detectable by all three
semantics.  This is a direct consequence of~\citet{gfd-oopsla-2019}'s
theorem that a run time type error under Erasure implies a run time type
error under both Transient and Natural.  It is worth noting that this
criterion excludes from consideration all scenarios where Natural or
Transient can detect an impedance but Erasure cannot, and thus it favors
Erasure.  However, our goal here is to test the rational programmer across
these three semantics, and thus we opt for a setting where all of them
notice the impedance.

The third criterion aims to capture debugging scenarios with a specific
kind of run time error: one produced by the interaction of a buggy
component with the rest of the program.  In those cases, the rational programmer
may need to examine several components to locate the source of the faulty interaction.
In technical terms, we conservatively mark these scenarios by
analyzing the stack trace of a run time error.  If the stack trace mentions at
least two distinct components, then this is evidence of interaction between
components.  We require a third component because our benchmarks come
with a driver component which is present in the stack trace of all run
time errors.

@@

With this definition in hand, we analyze how effective our mutators
are to turn the benchmarks into interesting debugging scenarios.
Indeed the mutators produce at least 72192 interesting debugging scenarios
across all mutants --- there are potentially many more but this is how
many we examined. 
Figure~\ref{fig:mutant-breakdown} breaks down the interesting
debugging scenarios produced  for each benchmark bu mutator 
but counts are cut off at 50. The mutators produce at least 40 interesting
scenarios for every benchmark, and these scenarios originate from at
least four different mutators per benchmark.  Thus the mutators result in 
a large and diverse population of scenarios for every benchmark.
Furthermore, every mutator contributes scenarios to at least one
benchmark and some mutators clearly only target one or two of them, which
justifies their inclusion. For
instance the class-focused mutators are primarily effective in
which jives with the fact that \texttt{take5} is the benchmark with 
the most extensive use of object-oriented features, while the other
features targeted by the rest of the mutators are used more broadly.

>>>>

However,  unlike the hard-to-kill mistakes introduced by standard mutators, 
gradual type systems flag mere typos.  Thus, in order to
exercise the rational programmer in the context of gradual typing, 
we need to develop alternative mutators
that are fine-tuned to produce type-level mistakes gradual type systems 
can detect. Our inspiration is our
more-than-a-decade long experience of making type-level mistakes in Typed
Racket. Some of these mistakes often take non-trivial effort to debug and
have become the seed for our mutators, which we summarize in
figure~\ref{table:mutation-ops}.  For each one, the figure provides a
short description and an example mutation.

Most of our mutators are self-explanatory.
The first twelve apply to most gradually typed languages, including those with classes.
The last four mutators target distinguishing features of Typed Racket.
Specifically, \texttt{arithmetic} may replace a \texttt{+} with a \texttt{-}
in an attempt to change the type of the result of the arithmetic
operation. In Typed Racket, \texttt{+}'s result is a
\texttt{Positive-Integer} when all arguments are
positive integers. However, the result of \texttt{-} is
\texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
advantage of Typed Racket's ``truthiness.'' Finally, \texttt{negate-cond} and \texttt{force-cond}
attempt to confuse occurrence typing.

@@

The astute reader must have observed that mutators that produce 
type-level mistakes do not necessarily turn our benchmarks into
interesting debugging scenarios for the rational programmer.  The
benchmarks are fully typed, and type checking detects each mistake at
compile time.  Hence, we apply the mutators to versions, dubbed
\dubbed{scenarios}, of the benchmarks where some components are
stripped of their type annotation.  Without the annotations, the type
checker may not detect a mistake at compile time and, instead, the run
time checks
of the gradual type system signal an error.  Scenarios
with type-level mistakes that result in run time errors are the ideal
ground to exercise the rational programmer's ability to locate errors.

Specifically, we define an \emph{interesting debugging scenario} as a
scenario that
\begin{enumerate} 
  \item is rejected by the type checker when we restore all its missing type annotations, 
  \item produces a run time error under Erasure as-is, and 
  \item the stack trace of the error mentions at least three distinct components.  
\end{enumerate}

The first criterion simply validates that the debugging scenario concerns
a type-level mistake. However, The second and third criteria require 
more explanation.  The second one stipulates that the mistake is
detectable under Erasure to ensure that it is detectable by all three
semantics.  This is a direct consequence of~\citet{gfd-oopsla-2019}'s
theorem that a run time error under Erasure implies a run time
error under both Transient and Natural.  It is worth noting that this
criterion excludes from consideration the very interesting scenarios where Natural or
Transient can detect a mistake but Erasure cannot, and thus it favors
Erasure.  However, our goal here is to test the rational programmer across
these three semantics, and thus we opt for a setting where all of them
notice the mistake.

The third criterion aims to capture debugging scenarios with a specific
kind of run time error: one produced by the interaction of a buggy
component with the rest of the program.  In those cases, the rational programmer
may need to examine several components to locate the source of the faulty interaction.
In technical terms, we mark these scenarios by
analyzing the stack trace of a run time error.  If the stack trace mentions at
least two distinct components, then this is accurate evidence of interaction between
components.  We require a third component because our benchmarks come
with a driver component which is present in the stack trace of all run
time errors.

@@

With this definition in hand, we analyze how effective our mutators
are for turning the benchmarks into interesting debugging scenarios.
Indeed the mutators produce at least 72192 interesting debugging scenarios
across all mutants --- there are potentially many more but this is how
many we examined. 
Figure~\ref{fig:mutant-breakdown} breaks down the interesting
debugging scenarios produced  for each benchmark by mutator 
but counts are cut off at 50. The mutators produce at least 40 interesting
scenarios for every benchmark, and these scenarios originate from at
least four different mutators per benchmark.  Thus the mutators result in 
a sizeable and diverse population of scenarios for every benchmark.
Furthermore, every mutator contributes scenarios to at least one
benchmark. Especially some mutators  target specific features of
benchmarks, which
justifies their inclusion. For
instance the class-focused mutators are effective in \texttt{take5}
which jives with the fact that \texttt{take5} is the benchmark with 
the most extensive use of object-oriented features.
